$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-05 paper/paper.tex
Pass to intro, contrib, syntax, semantics, and contexts

Let $x, y,\ldots$ range over \emph{variables} (including channel names) and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

@@

We define the base cases of a structural congruence relation \textsc{StructCong} $\eqeq$ in \autoref{fig:struct-cong-base}.
Its congruent equivalence closure \textsc{Equals} $=$ is defined1 in \autoref{app:struct}.

@@

The operational semantics of the \picalc{} is defined by a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{subject-reduction}).
We identify two kinds of channels, $\constr{internal}$ and $\constr{external}\ i$, respectively standing for a channel which ... \todo{complete}

@@

We use a $\func{dec}$ function to decrement the index of this channel every time we come out of a binder in rule $\constr{res}$. \todo{this seems more getting 'into' a binder than coming out...}
This $\func{dec}$ function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, which is \emph{partial}, \emph{functional}, \emph{cancellative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} uses a terniary relation to model such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

    \item [Gradual]
    The carrier is the type of natural numbers.
    The element $\zero$ corresponds to 0, and $\one$ to 1.
    The partial monoid $\op{x}{y}{z}$ is defined exactly when $x \equiv y + z$. \todo{why isnt $\equiv$ here just =?}
    All other properties of the algebra follow from the algebraic rules for the addition of natural numbers.
    As an example, we use $(5 \comma 3)$ to represent the multiplicities of a channel that is able to input 5 times and output 3 times.

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[IDX set of usage algebra]
  An \emph{indexed set of usage algebras} is a nonempty type $\Idx$ of indices together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation}
      \begin{split}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{split}
    \end{equation}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

  We keep track of types in a \emph{length-indexed typing context} $\PreCtx_n$, defined by the last row of rules in \autoref{fig:prectx}.
  A typing context is either empty $[]$ or building upon an existing typing context $\gamma$, it associates a type $t$ to an the next index in the context.
  \todo{check the last sentence}
  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : \Set}
    \; \textsc{Type}

@@

\begin{nidefinition}[Usage context]
  A \emph{usage context} in a context $\Ctx_{idxs}$ that is indexed by a length-indexed context of indices $\Idxs_n$, where we keep track of usage annotations (\autoref{fig:ctx}).

  A usage context is either empty $[]$ or building upon an existing context $\Gamma$, it associates a usage to the current index.
  \todo{check the last sentence}

>>>>

\todo{dots}
Let $x, y,\ldots$ range over \emph{variables} (including channel names) and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

@@

We define the base cases of a structural congruence relation \textsc{StructCong} $\eqeq$ in \autoref{fig:struct-cong-base}.
Its congruent equivalence closure \textsc{Equals} $\eq{}$ is defined in \autoref{app:struct}.

@@

The operational semantics of the \picalc{} is defined by a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{subject-reduction}).
We distinguish between channels created inside a process ($\constr{internal}$) and channels created outside ($\constr{external}\ i$, where $i$ is the de Bruijn index of the variable).

@@

In $\constr{res}$, we use a $\func{dec}$ function to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
This $\func{dec}$ function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, that is \emph{partial}, \emph{functional}, \emph{cancellative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} uses a terniary relation to model such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

    \item [Gradual]
    The carrier is the type of natural numbers.
    The element $\zero$ corresponds to 0, and $\one$ to 1.
    The partial monoid $\op{x}{y}{z}$ is defined exactly when $x = y + z$.
    All other properties of the algebra follow from the algebraic rules for the addition of natural numbers.
    As an example, we use $(5 \comma 3)$ to represent the multiplicities of a channel that is able to input 5 times and output 3 times.

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a nonempty type $\Idx$ of indices together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation}
      \begin{split}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{split}
    \end{equation}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

  We keep track of types in a \emph{length-indexed typing context} $\PreCtx_n$, defined by the last row of rules in \autoref{fig:prectx}.
  A typing context is either empty $[]$ or the result of appending a type $t$ to an existing context$\gamma$.
  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : \Set}
    \; \textsc{Type}

@@

\begin{nidefinition}[Usage contexts]
  A \emph{usage context} is a context $\Ctx_{idxs}$ of usage annotations that is indexed by a length-indexed context of indices $\Idxs_n$ (\autoref{fig:ctx}).

  A usage context is either empty $[]$ or the result or appending a usage annotation $x$ to the existing context $\Gamma$.
